---
title: "COMP7024: Programming for Data Science"
subtitle: "Assignment Quarter 1 2026"
author: "Harriet Nafula Wanyonyi Student ID: 22119241"
date: "2026-03-10"
output: pdf_document
---

## Project Cover Page & Declaration

By including this statement, we the authors of this work, verify that:
* We hold a copy of this assignment that we can produce if the original is lost or damaged.
* We hereby certify that no part of this assignment/product has been copied from any other student's work or from any other source except where due acknowledgement is made in the assignment.
* No part of this assignment/product has been written/produced for us by another person except where such collaboration has been authorised by the subject lecturer/tutor concerned.
* We are aware that this work may be reproduced and submitted to plagiarism detection software programs for the purpose of detecting possible plagiarism.
* We hereby certify that we have read and understand what the School of Computing, Engineering and Mathematics defines as minor and substantial breaches of misconduct.

\newpage

## 1. Data Preparation & Product Revenue Analysis

### Task Explanation & Assumptions

This task requires combining our transactional sales records with the product information dataset to attach descriptions to each transaction. Before conducting any revenue calculations, we must address data quality by filtering out cancelled orders. 

Our core processing rationales and assumptions are:
1. **Relational Merging:** We apply an inner join to link the transactional data with the product dictionary using the unique `StockCode` identifier. This ensures every analyzed row contains complete product names.
2. **Cancellations:** Transaction records with an `InvoiceNo` beginning with the letter "C" denote cancelled transactions. These must be removed completely to avoid artificial skewing of sales volumes.
3. **Revenue Definition:** Financial performance is tracked by calculating gross revenue as the product of the transactional `Quantity` and its specific `UnitPrice`.

### R Code & Analysis

```{r task1_analysis, message=FALSE, warning=FALSE}
# Load core tidyverse packages and layout tools
library(tidyverse)
library(knitr)

# Load the raw transactional datasets
orders <- read_csv("project_orders.csv")
product_info <- read_csv("project_product_info.csv")

# Filter out cancellations and execute the mandatory relational join
clean_orders <- orders %>%
  filter(!str_detect(InvoiceNo, "^C")) %>%
  inner_join(product_info, by = "StockCode")

# Aggregate metrics at the product profile level
product_summary <- clean_orders %>%
  group_by(StockCode, Description) %>%
  summarise(
    Total_Quantity = sum(Quantity),
    Total_Revenue = sum(Quantity * UnitPrice),
    .groups = "drop"
  )

# Isolate the top 10 highest-performing products by total revenue
top_10_products <- product_summary %>%
  slice_max(order_by = Total_Revenue, n = 10)

# Isolate the bottom 10 lowest-performing products by total revenue
bottom_10_products <- product_summary %>%
  slice_min(order_by = Total_Revenue, n = 10)

# Evaluate revenue concentration for the top 3 products
global_revenue_total <- sum(product_summary$Total_Revenue)

top_3_concentration <- product_summary %>%
  slice_max(order_by = Total_Revenue, n = 3) %>%
  summarise(
    Combined_Top3_Revenue = sum(Total_Revenue),
    Global_Revenue = global_revenue_total,
    Percentage_Share = (Combined_Top3_Revenue / global_revenue_total) * 100
  )

# Output well-formatted markdown tables for the final PDF document
kable(top_10_products, caption = "Top 10 Highest-Selling Products by Revenue")
kable(bottom_10_products, caption = "Bottom 10 Lowest-Selling Products by Revenue")
kable(top_3_concentration, caption = "Top 3 Products Revenue Concentration Summary")
```


## Code Testing & Verification

```{r task1_testing}
# Ensure that absolutely zero cancelled invoices remain in cleaned dataset
cancelled_count <- sum(str_detect(clean_orders$InvoiceNo, "^C"))
print(paste("Number of remaining cancelled transactions:", cancelled_count))

# Verify that no row contains missing product descriptions post-merge
missing_desc_count <- sum(is.na(clean_orders$Description))
print(paste("Number of records missing descriptions:", missing_desc_count))
```
