---
title: "COMP7024: Programming for Data Science"
subtitle: "Assignment Quarter 1 2026"
author: "Harriet Nafula Wanyonyi Student ID: 22119241"
date: "2026-03-10"
output: pdf_document
---

## Project Cover Page & Declaration

By including this statement, we the authors of this work, verify that:
* We hold a copy of this assignment that we can produce if the original is lost or damaged.
* We hereby certify that no part of this assignment/product has been copied from any other student's work or from any other source except where due acknowledgement is made in the assignment.
* No part of this assignment/product has been written/produced for us by another person except where such collaboration has been authorised by the subject lecturer/tutor concerned.
* We are aware that this work may be reproduced and submitted to plagiarism detection software programs for the purpose of detecting possible plagiarism.
* We hereby certify that we have read and understand what the School of Computing, Engineering and Mathematics defines as minor and substantial breaches of misconduct.

\newpage

## 1. Data Preparation & Product Revenue Analysis

### Task Explanation & Assumptions

This task requires combining our transactional sales records with the product information dataset to attach descriptions to each transaction. Before conducting any revenue calculations, we must address data quality by filtering out cancelled orders. 

Our core processing rationales and assumptions are:
1. **Relational Merging:** We apply an inner join to link the transactional data with the product dictionary using the unique `StockCode` identifier. This ensures every analyzed row contains complete product names.
2. **Cancellations:** Transaction records with an `InvoiceNo` beginning with the letter "C" denote cancelled transactions. These must be removed completely to avoid artificial skewing of sales volumes.
3. **Revenue Definition:** Financial performance is tracked by calculating gross revenue as the product of the transactional `Quantity` and its specific `UnitPrice`.

### R Code & Analysis

```{r task1_analysis, message=FALSE, warning=FALSE}
# Load core tidyverse packages and layout tools
library(tidyverse)
library(knitr)

# Load the raw transactional datasets
orders <- read_csv("project_orders.csv")
product_info <- read_csv("project_product_info.csv")

# Filter out cancellations and execute the mandatory relational join
clean_orders <- orders %>%
  filter(!str_detect(InvoiceNo, "^C")) %>%
  inner_join(product_info, by = "StockCode")

# Aggregate metrics at the product profile level
product_summary <- clean_orders %>%
  group_by(StockCode, Description) %>%
  summarise(
    Total_Quantity = sum(Quantity),
    Total_Revenue = sum(Quantity * UnitPrice),
    .groups = "drop"
  )

# Isolate the top 10 highest-performing products by total revenue
top_10_products <- product_summary %>%
  slice_max(order_by = Total_Revenue, n = 10)

# Isolate the bottom 10 lowest-performing products by total revenue
bottom_10_products <- product_summary %>%
  slice_min(order_by = Total_Revenue, n = 10)

# Evaluate revenue concentration for the top 3 products
global_revenue_total <- sum(product_summary$Total_Revenue)

top_3_concentration <- product_summary %>%
  slice_max(order_by = Total_Revenue, n = 3) %>%
  summarise(
    Combined_Top3_Revenue = sum(Total_Revenue),
    Global_Revenue = global_revenue_total,
    Percentage_Share = (Combined_Top3_Revenue / global_revenue_total) * 100
  )

# Output well-formatted markdown tables for the final PDF document
kable(top_10_products, caption = "Top 10 Highest-Selling Products by Revenue")
kable(bottom_10_products, caption = "Bottom 10 Lowest-Selling Products by Revenue")
kable(top_3_concentration, caption = "Top 3 Products Revenue Concentration Summary")
```


## Code Testing & Verification

```{r task1_testing}
# Ensure that absolutely zero cancelled invoices remain in cleaned dataset
cancelled_count <- sum(str_detect(clean_orders$InvoiceNo, "^C"))
print(paste("Number of remaining cancelled transactions:", cancelled_count))

# Verify that no row contains missing product descriptions post-merge
missing_desc_count <- sum(is.na(clean_orders$Description))
print(paste("Number of records missing descriptions:", missing_desc_count))
```


## Data Visualisation

Below is a horizontal bar chart of our top 10 inventory items to visualize sales performance distribution.

```{r task1_plot, fig.height=5, fig.width=8}
# Construct a sorted horizontal bar chart for clear interpretation
ggplot(top_10_products, aes(x = reorder(Description, Total_Revenue), y = Total_Revenue)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Top 10 Products by Total Revenue",
    subtitle = "Excludes cancelled transactions",
    x = "Product Description",
    y = "Total Revenue (£)"
  ) +
  theme_minimal()
```


## Interpretation of Results

The primary metrics show that the remaining key inventory items command less performance focus. Top of the list with £164,814.20 of sales is the REGENCY CAKESTAND 3 TIER. The next two items earned less than 5.49% of platform sales combined. This shows a good baseline of demand for the items, and the platform does not depend on narrower scope sales and revenue to function.

On the contrary, the inventory items that are lowest in sales have negative monetary balances. This includes the item 'Adjust Bad Debt' (-£11,062.06) and various other handling designations. In the normal running of an organization, these records represent financial adjustments, clearance write-offs and adjustments of inventory write-offs due to damages. These are not normal retail transactions. From a management point of view, the presence of negative balances is a concern. The focus should be to control these balances and mitigate exposure of the inventory assets of the organization.

## 2. High-Value Customer Analysis

### Explanation & Assumptions

This project analyzes transactional data to find our best customers. Specifically, we find the top 5 customers who have contributed the most total revenue. After that, we do a deep dive into the data to find their most commonly purchased items by total quantity.

Our core processing rationales and assumptions are:
1. **Valid Customer Tracking:** Any rows missing a valid CustomerID are excluded from the ranking calculation, as anonymous transactions cannot be attributed to specific high-value customer retention strategies.
2. **Revenue Metrics Alignment:** Customer spending rankings are determined using the clean data from Task 1, which excludes cancelled orders.
3. **Volume Definition:** The product preferences for these top buyers are ranked based on the sum of the Quantity variable for each distinct product.

### Code

```{r task2_analysis, message=FALSE, warning=FALSE}
# Identify the top 5 highest-spending customers
top_5_customers <- clean_orders %>%
  filter(!is.na(CustomerID)) %>%
  group_by(CustomerID) %>%
  summarise(Total_Spent = sum(Quantity * UnitPrice), .groups = "drop") %>%
  slice_max(order_by = Total_Spent, n = 5)

# Filter the main dataset to isolate all records for these 5 customers
top_customer_orders <- clean_orders %>%
  filter(CustomerID %in% top_5_customers$CustomerID)

# Identify the most frequently purchased product for each of the top 5 customers
top_products_per_customer <- top_customer_orders %>%
  group_by(CustomerID, StockCode, Description) %>%
  summarise(Total_Quantity = sum(Quantity), .groups = "drop") %>%
  group_by(CustomerID) %>%
  slice_max(order_by = Total_Quantity, n = 1) %>%
  ungroup()

# Output clean tabular formats for the report layout
knitr::kable(top_5_customers, caption = "Top 5 Highest Spending Customers by Total Revenue")
knitr::kable(top_products_per_customer, caption = "Most Frequently Purchased Product by Volume for Top Customers")
```


### Code Testing & Verification

```{r task2_testing}
# Ensure exactly 5 unique high-value customers are being analyzed
unique_customer_count <- length(unique(top_5_customers$CustomerID))
print(paste("Number of unique high-value customers isolated:", unique_customer_count))

# Confirm that no missing Customer IDs bypassed our filter stage
missing_ids <- sum(is.na(top_customer_orders$CustomerID))
print(paste("Number of missing customer records in filtered cohort:", missing_ids))
```


### Data Visualisation
To find the large quantity purchasing habits of our customers, we create a faceted bar chart to show the total quantity of items customers purchased.

```{r task2_plot, fig.height=6, fig.width=10}
# Convert CustomerID to a discrete factor to remove scientific notation
# Wrap long descriptions so they do not overlap or cut off
plot_data <- top_products_per_customer %>%
  mutate(
    CustomerID = as.factor(CustomerID),
    Clean_Description = stringr::str_wrap(Description, width = 25)
  )

# Create a grouped visualization with clean wrapped text labels
ggplot(plot_data, aes(x = reorder(Clean_Description, Total_Quantity), y = Total_Quantity, fill = CustomerID)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~CustomerID, scales = "free") +
  coord_flip() +
  labs(
    title = "Top Product Volumes for High-Value Customers",
    subtitle = "Isolates the highest-volume item per customer profile",
    x = "Product Description",
    y = "Total Physical Volume Purchased"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 9, lineheight = 0.8),
    strip.text = element_text(face = "bold", size = 11)
  )
```

### Interpretation of Results

This part of our analysis shows the top 5 customers who spent the most. The customer who has spent the most is Customer ID 14646 with total spending of £280,206.02, while Customer ID 18102 shows the second highest spending at £259,657.30. The top 5 customers are each representative of a different large customer buying wholesale from our organization.

Using our faceted visualization, we see that each of the three customers has unique purchasing habits. For example, Customer ID 14646 has purchased more than 4,000 of the item 'RABBIT NIGHT LIGHT', while Customer ID 12415 has purchased more than 4,000 of the item 'PACK OF 72 RETROSPOT CAKE CASES'. Each of these customers contributes large amounts of revenue to our organization, and we should apply special pricing policies to them to ensure that they stay customers.



## Task 3: Global Market Analysis

### Explanation & Assumptions

This task investigates the geographic reach of the retailer's customer base. We consolidate financial data to identify how much revenue each country contributes. Additionally, we examine each country’s transaction data to determine its leading product category by total physical sales volume (Quantity).

The reasoned assumptions that form the basis of our processing are:

1. *Geographic Inclusion*: All transactions with a valid country identifier are included. Clean data from Task 1 is used to maintain consistency in calculations and to ensure excluded order cancellations.
2. *Popularity Measure*: Within each country, product popularity is measured exclusively by physical volume, which is determined by summing the Quantity variable of each unique stock item.
3. *Tie-Breaker Policy*: When multiple products in a given country exhibit the same maximum quantity, the `slice_max(..., n = 1)` function will select the first matching product to maintain simplicity in the summary table.

### Code

```{r task3_analysis, message=FALSE, warning=FALSE}
# Geographic Aggregation: Calculate total revenue per country
country_revenue <- clean_orders %>%
  group_by(Country) %>%
  summarise(Total_Revenue = sum(Quantity * UnitPrice), .groups = "drop") %>%
  arrange(desc(Total_Revenue))

# Identify the most popular product by volume for each country
# We use str_wrap directly in the data prep so the table and plot labels stay clean!
country_top_products <- clean_orders %>%
  group_by(Country, StockCode, Description) %>%
  summarise(Total_Quantity = sum(Quantity), .groups = "drop") %>%
  group_by(Country) %>%
  slice_max(order_by = Total_Quantity, n = 1) %>%
  ungroup() %>%
  mutate(Clean_Description = stringr::str_wrap(Description, width = 25)) %>%
  arrange(desc(Total_Quantity))

# Output clean tabular formats for the top 10 rows for the report layout
knitr::kable(head(country_revenue, 10), caption = "Top 10 Countries by Total Revenue Generation")
knitr::kable(head(country_top_products[, c("Country", "StockCode", "Description", "Total_Quantity")], 10), 
             caption = "Most Popular Product by Volume Across Top 10 Countries")
```


### Code Testing & Verification

```{r task2_testing}
# Verify that every country present in the clean dataset has exactly one top product entry
total_unique_countries <- length(unique(clean_orders$Country))
top_products_rows <- nrow(country_top_products)
print(paste("Total unique countries:", total_unique_countries))
print(paste("Number of rows in country top products table:", top_products_rows))

# Confirm that the total calculated country revenue matches our global baseline revenue
global_baseline <- sum(clean_orders$Quantity * clean_orders$UnitPrice)
summed_country_revenue <- sum(country_revenue$Total_Revenue)
print(paste("Does summed country revenue match global baseline?:", identical(all.equal(global_baseline, summed_country_revenue), TRUE)))
```


### Data Visualisation

To analyze performance by country, we develop a horizontal bar chart that shows the revenue total for each of the top international markets. We omit the primary domestic market (United Kingdom) to ensure that the remaining countries displayed on the chart are scaled legibly and appropriately.


```{r task3_plot, message=FALSE, warning=FALSE, fig.height=5, fig.width=9}
# Load libraries inside the chunk to secure absolute execution independence
library(tidyverse)

# Isolate top 10 international markets excluding the United Kingdom for clean scaling
top_international <- country_revenue %>%
  filter(Country != "United Kingdom") %>%
  slice_max(order_by = Total_Revenue, n = 10)

# Build the publication-ready bar chart
ggplot(top_international, aes(x = reorder(Country, Total_Revenue), y = Total_Revenue)) +
  geom_col(fill = "darkred") +
  coord_flip() +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Top 10 International Markets by Revenue",
    subtitle = "Excludes the United Kingdom to prevent visual scaling distortion",
    x = "Country",
    y = "Total Revenue (£)"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 10, face = "bold"),
    axis.text.x = element_text(size = 9)
  )
```


### Interpretation of Results

The geographic revenue analysis reveals that the domestic market is our primary driver of financial performance, with the United Kingdom contributing an overwhelming £9,003,097.96 to overall operations. Outside the domestic market, the Netherlands leads international sales with £285,446.34, followed closely by EIRE at £283,453.96 and Germany at £228,867.14. This strong clustering among Western European nations suggests that regional proximity, logistical simplicity, and existing marketing campaigns are highly effective in these territories.

When assessing localized product preferences, the data shows distinct regional commonalities. For example, the Netherlands, France, and Japan all share a common favorite item by physical volume: the 'RABBIT NIGHT LIGHT', which achieved peak volumes of 4,801 units in the Netherlands alone. Meanwhile, the domestic UK market shows a unique preference for bulk volumes of 'PAPER CRAFT, LITTLE BIRDIE' at 80,995 units. From a supply chain perspective, these insights imply that inventory management should be regionalized. High-volume hubs like the Netherlands can be utilized to optimize shipping for cross-border top-sellers, minimizing fulfillment delays for our prominent international markets.


## Task 4: Product Recommendation System


### Explanation & Assumptions

In this task, we create a product recommendation engine using market basket analysis. The required product is a function that takes the StockCode of a target product and finds which other products are purchased most frequently with it on the same invoice.

Here are the main ideas and assumptions that drive our processing.

1. **Transaction Co-occurrence:** When two products are recorded on the same InvoiceNo, we consider them co-purchased.
2. **Self-Exclusion Rule:** We do not include the target product in the recommendations since that would create an infinite loop.
3. **Clean Catalog References:** We pull product descriptions from the product info dataset to clarify the output for end users.


### Code

```{r task4_analysis, message=FALSE, warning=FALSE}
# Load required libraries within the chunk to ensure complete independence
library(tidyverse)

# Define the standalone recommendation function
get_product_recommendations <- function(target_stock_code, top_n = 3) {
  
  # Identify all unique invoices that contain the target product
  target_invoices <- clean_orders %>%
    filter(StockCode == target_stock_code) %>%
    pull(InvoiceNo) %>%
    unique()
  
  # Isolate all other products present within those exact invoices
  recommendations <- clean_orders %>%
    filter(InvoiceNo %in% target_invoices & StockCode != target_stock_code) %>%
    group_by(StockCode, Description) %>%
    summarise(Co_Purchase_Count = n(), .groups = "drop") %>%
    slice_max(order_by = Co_Purchase_Count, n = top_n, with_ties = FALSE)
  
  return(recommendations)
}

# Run a test execution using our top-performing item: REGENCY CAKESTAND 3 TIER (StockCode: 22423)
test_recommendations <- get_product_recommendations(target_stock_code = "22423", top_n = 3)

# Output a clean tabular format for the report layout
knitr::kable(test_recommendations, caption = "Top 3 Product Recommendations for StockCode 22423 (REGENCY CAKESTAND 3 TIER)")
```


### Testing & Verification

```{r task4_testing&verification, message=FALSE, warning=FALSE}
library(tidyverse)
# Ensure the function returns exactly the number of products requested (3)
returned_rows <- nrow(test_recommendations)
print(paste("Number of product recommendations generated:", returned_rows))
#Confirm that the target item itself was successfully excluded from the outputs
self_matching_records <- sum(test_recommendations$StockCode == "22423")
print(paste("Number of self-matching entries in output table:", self_matching_records))
```


### Interpretation of Results

The product recommendation system has employed cross-transactional history to discover latent links between consumer purchases. Analyzing historical transactions for one of our top sellers, the 'REGENCY CAKESTAND 3 TIER' (StockCode: 22423), allows the system to detect items on the same invoice. Results indicate a strong cross-selling relationship for a single themed product group. Based on the results, 'ROSES REGENCY TEACUP AND SAUCER' tops the recommendation with 537 purchases. It is followed by 'GREEN REGENCY TEACUP AND SAUCER' with 529 purchases, and 'PINK REGENCY TEACUP AND SAUCER' with 416 purchases.

These results are very informative from commercial and operational perspectives for optimizing the value of transactions. Rather than viewing these items as separate product catalog items, marketing teams should sell these items together as a cohesive product collection or offer bundled promotions. Additionally, the web design team should use these data points and put a 'Complete the Look' or 'Frequently Bought Together' along with the line of products on the product description page to facilitate user navigation and cross-purchase. This implementation should generate an increase in the average basket value across the retail operations.



