---
title: "COMP7024: Programming for Data Science"
subtitle: "Assignment Quarter 1 2026"
author: "Harriet Nafula Wanyonyi Student ID: 22119241"
date: "2026-03-10"
output: pdf_document
---

## Project Cover Page & Declaration

By including this statement, we the authors of this work, verify that:
* We hold a copy of this assignment that we can produce if the original is lost or damaged.
* We hereby certify that no part of this assignment/product has been copied from any other student's work or from any other source except where due acknowledgement is made in the assignment.
* No part of this assignment/product has been written/produced for us by another person except where such collaboration has been authorised by the subject lecturer/tutor concerned.
* We are aware that this work may be reproduced and submitted to plagiarism detection software programs for the purpose of detecting possible plagiarism.
* We hereby certify that we have read and understand what the School of Computing, Engineering and Mathematics defines as minor and substantial breaches of misconduct.

\newpage

## 1. Data Preparation & Product Revenue Analysis

### Task Explanation & Assumptions

This task requires combining our transactional sales records with the product information dataset to attach descriptions to each transaction. Before conducting any revenue calculations, we must address data quality by filtering out cancelled orders. 

Our core processing rationales and assumptions are:
1. **Relational Merging:** We apply an inner join to link the transactional data with the product dictionary using the unique `StockCode` identifier. This ensures every analyzed row contains complete product names.
2. **Cancellations:** Transaction records with an `InvoiceNo` beginning with the letter "C" denote cancelled transactions. These must be removed completely to avoid artificial skewing of sales volumes.
3. **Revenue Definition:** Financial performance is tracked by calculating gross revenue as the product of the transactional `Quantity` and its specific `UnitPrice`.

### R Code & Analysis

```{r task1_analysis, message=FALSE, warning=FALSE}
# Load core tidyverse packages and layout tools
library(tidyverse)
library(knitr)

# Load the raw transactional datasets
orders <- read_csv("project_orders.csv")
product_info <- read_csv("project_product_info.csv")

# Filter out cancellations and execute the mandatory relational join
clean_orders <- orders %>%
  filter(!str_detect(InvoiceNo, "^C")) %>%
  inner_join(product_info, by = "StockCode")

# Aggregate metrics at the product profile level
product_summary <- clean_orders %>%
  group_by(StockCode, Description) %>%
  summarise(
    Total_Quantity = sum(Quantity),
    Total_Revenue = sum(Quantity * UnitPrice),
    .groups = "drop"
  )

# Isolate the top 10 highest-performing products by total revenue
top_10_products <- product_summary %>%
  slice_max(order_by = Total_Revenue, n = 10)

# Isolate the bottom 10 lowest-performing products by total revenue
bottom_10_products <- product_summary %>%
  slice_min(order_by = Total_Revenue, n = 10)

# Evaluate revenue concentration for the top 3 products
global_revenue_total <- sum(product_summary$Total_Revenue)

top_3_concentration <- product_summary %>%
  slice_max(order_by = Total_Revenue, n = 3) %>%
  summarise(
    Combined_Top3_Revenue = sum(Total_Revenue),
    Global_Revenue = global_revenue_total,
    Percentage_Share = (Combined_Top3_Revenue / global_revenue_total) * 100
  )

# Output well-formatted markdown tables for the final PDF document
kable(top_10_products, caption = "Top 10 Highest-Selling Products by Revenue")
kable(bottom_10_products, caption = "Bottom 10 Lowest-Selling Products by Revenue")
kable(top_3_concentration, caption = "Top 3 Products Revenue Concentration Summary")
```


## Code Testing & Verification

```{r task1_testing}
# Ensure that absolutely zero cancelled invoices remain in cleaned dataset
cancelled_count <- sum(str_detect(clean_orders$InvoiceNo, "^C"))
print(paste("Number of remaining cancelled transactions:", cancelled_count))

# Verify that no row contains missing product descriptions post-merge
missing_desc_count <- sum(is.na(clean_orders$Description))
print(paste("Number of records missing descriptions:", missing_desc_count))
```


## Data Visualisation

Below is a horizontal bar chart of our top 10 inventory items to visualize sales performance distribution.

```{r task1_plot, fig.height=5, fig.width=8}
# Construct a sorted horizontal bar chart for clear interpretation
ggplot(top_10_products, aes(x = reorder(Description, Total_Revenue), y = Total_Revenue)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Top 10 Products by Total Revenue",
    subtitle = "Excludes cancelled transactions",
    x = "Product Description",
    y = "Total Revenue (£)"
  ) +
  theme_minimal()
```


## Interpretation of Results

The primary metrics show that the remaining key inventory items command less performance focus. Top of the list with £164,814.20 of sales is the REGENCY CAKESTAND 3 TIER. The next two items earned less than 5.49% of platform sales combined. This shows a good baseline of demand for the items, and the platform does not depend on narrower scope sales and revenue to function.

On the contrary, the inventory items that are lowest in sales have negative monetary balances. This includes the item 'Adjust Bad Debt' (-£11,062.06) and various other handling designations. In the normal running of an organization, these records represent financial adjustments, clearance write-offs and adjustments of inventory write-offs due to damages. These are not normal retail transactions. From a management point of view, the presence of negative balances is a concern. The focus should be to control these balances and mitigate exposure of the inventory assets of the organization.
